Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

### 2024-06-15 EBDVF-2024/paper.qmd
Update paper.qmd

---
title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
author:
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    affiliations: 
        - id: swc
          name: The Semantic Web Company
          country: Austria
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliations: 
        - id: onto
          name: Ontotext
          country: Bulgaria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
    affiliations:
        - ref: onto
abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
  The Vocabulary Hub service (part of the IDSA RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.
keywords: 
  - dataspaces
  - semantic interoperability
  - semantic technologies
  - ontologies
  - vocabulary hub
  - oil and gas
  - renewable energy
  - refineries
  - windfarms
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Workshop on Semantic Interoperability in Data Spaces 
      formatting: preprint
      model: 3p
      cite-style: super
---

@@

## IDS Vocabulary Hub
In this paper, we focus on the IDSA approach to dataspaces. 
IDSA defines the architecture of dataspaces in the reference architecture model IDS-RAM (current version 4). 
One of the defined architectural components, the Vocabulary Hub, has the role of providing vocabularies to annotate and describe data assets and services. 
These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions. 
It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies 
and thereby provides semantic interoperability in a standardized and machine-readable way.

The Vocabulary Hub is one of the core components of any dataspace not only those designed in accordance with IDS-RAM. 
Other reference architecture models, such as GAIA-X architecture document 22.10 Release [@GaiaXArchitectureDocument2022] and Fraunhofer's Reference Architecture Model for the Industrial Data Space [@ReferenceArhictectureModel2017] also describe components with a similar role - the Federated Catalogue and Broker Service, respectively. 
Therefore, our approach designed around the IDS-RAM is adaptable and applicable to data spaces designed in accordance with other reference architecture models.

@@

## Semantic Layer Approach
The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
Those vocabularies are to be expressed in RDF and by syntactically compatible with OWL to allow for expressive semantics. 
The IDS-RAM does not require, but nonetheless allows for, extending the function of the Vocabulary hub towards providing ontology mappings [@FunctionalLayerIDSRAM2023b], which would enable dataspace connectors to automatically convert data between user-specific and standard data formats. Such automatic conversions are to be performed by Data Tranformation Apps by either the Data Consumers, the Data Providers, or both.

@@

- Bringing it into actual dataspaces in practice, representing and integrating datasets from different dataspace participants.
- Implementing use cases for richer discoverability, harmonized querying and support for different content types of structured and unstructured data.
- Explore how ML can benefit regarding quality in practice when providing consolidated and cleaned data via dataspaces.
- Discuss how we can extend the IDS-RAM with services to improve the support for semantic interoperability provided by our solution.

>>>>

---
title: Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces
author:
  - name: Robert David
    email: robert.david@semantic-web.com
    orcid: 0000-0002-3244-5341
    affiliations: 
        - id: swc
          name: The Semantic Web Company
          country: Austria
  - name: Petar Ivanov
    email: petar.ivanov@ontotext.com
    orcid: 0000-0001-8448-1005
    affiliations: 
        - id: onto
          name: Ontotext
          country: Bulgaria
  - name: Vladimir Alexiev
    email: vladimir.alexiev@ontotext.com
    orcid: 0000-0001-7508-7428
    affiliations:
        - ref: onto
abstract: |
  Dataspaces are an important enabler for industrial sharing data (either commercially licensed or private).
  Europe is investing heavily into sectoral dataspaces, federation and orchestration platforms like SIMPL, Eclipse DSC, GXFS, etc.
  Still, dataspaces enable shared data access, but do not solve the data interoperability problem.
  For that, the consumer would like to see the data from different providers in a harmonized and semantically integrated form.
  The Vocabulary Hub service (part of the IDS RAM) provides a repository for ontologies and vocabularies.
  We describe an approach of raising the role of the vocabulary hub to also allow richer metadata description (e.g. the meaning of every column in a tabular dataset),
  and binding semantic descriptions to ingested datasets, thus providing on-the-fly data semantization and easing data querying.
  This is achieved through the integration of two commercial semantic products (PoolParty and GraphDB),
  leveraging the partnership between the Semantic Web Company and Ontotext,
  and is being developed within the frame of the Digital Europe project UNDERPIN,
  with applications to refinery and wind farm data.
keywords: 
  - dataspaces
  - semantic interoperability
  - semantic technologies
  - ontologies
  - vocabulary hub
  - oil and gas
  - renewable energy
  - refineries
  - windfarms
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: Workshop on Semantic Interoperability in Data Spaces 
      formatting: preprint
      model: 3p
      cite-style: super
---

@@

## IDS Vocabulary Hub
In this paper, we focus on the IDSA approach to dataspaces. 
IDSA defines the architecture of dataspaces in the reference architecture model IDS RAM (current version 4). 
One of the defined architectural components, the Vocabulary Hub, has the role of providing vocabularies to annotate and describe data assets and services. 
These vocabularies form a common language to enable semantic interoperability among the dataspace participants. 
The Vocabulary Hub is responsible for resolving annotations of data assets and for providing details in the form of standardized semantic descriptions. 
It supports the use of Semantic Web standards for defining and representing RDF vocabularies and ontologies 
and thereby provides semantic interoperability in a standardized and machine-readable way.

The Vocabulary Hub is one of the core components of any dataspace not only those designed in accordance with IDS RAM. 
Other reference architecture models, such as GAIA-X architecture document 22.10 Release [@GaiaXArchitectureDocument2022] and Fraunhofer's Reference Architecture Model for the Industrial Data Space [@ReferenceArhictectureModel2017] also describe components with a similar role - the Federated Catalogue and Broker Service, respectively. 
Therefore, our approach designed around the IDS RAM is adaptable and applicable to data spaces designed in accordance with other reference architecture models.

@@

## Semantic Layer Approach
The Vocabulary Hub, as IDS defines it in the IDS RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
Those vocabularies are to be expressed in RDF and by syntactically compatible with OWL to allow for expressive semantics. 
The IDS RAM does not require, but nonetheless allows for, extending the function of the Vocabulary hub towards providing ontology mappings [@FunctionalLayerIDSRAM2023b], which would enable dataspace connectors to automatically convert data between user-specific and standard data formats. Such automatic conversions are to be performed by Data Tranformation Apps by either the Data Consumers, the Data Providers, or both.

@@

- Bringing it into actual dataspaces in practice, representing and integrating datasets from different dataspace participants.
- Implementing use cases for richer discoverability, harmonized querying and support for different content types of structured and unstructured data.
- Explore how ML can benefit regarding quality in practice when providing consolidated and cleaned data via dataspaces.
- Discuss how we can extend the IDS RAM with services to improve the support for semantic interoperability provided by our solution.
